---
title: "ggparty"
output: rmarkdown::github_document
---

**ggplot2** visualizations for the **partykit** package.

## Install 
```{r, eval=FALSE}
devtools::install_github("martin_borkovec/ggparty", 
                         dependencies = TRUE)
```

## Example
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dpi = 300,
                      fig.path = "man/figures/README-")
options(warn = -1)
```



```{r, warning = FALSE, message = FALSE, fig.width = 15, fig.height = 10}
library(ggparty)

data("TeachingRatings", package = "AER")
tr <- subset(TeachingRatings, credits == "more")

tr_tree <- lmtree(eval ~ beauty | minority + age + gender + division + native +
                    tenure, data = tr, weights = students, caseweights = FALSE)

ggparty(tr_tree,
        terminal_space = 0.5,
        add_vars = list(p.value = "$node$info$p.value")) +
  geom_edge(size = 1.5) +
  geom_edge_label(colour = "grey", size = 6) +
  geom_node_plot(gglist = list(geom_point(aes(x = beauty,
                                             y = eval,
                                             col = tenure,
                                             shape = minority),
                                         alpha = 0.8),
                              theme_bw(base_size = 15)),
                scales = "fixed",
                id = "terminal",
                shared_axis_labels = T,
                shared_legend = T,
                legend_separator = T,
                predict = "beauty",
                predict_gpar = list(col = "blue",
                                   size = 1.2)
                ) +
  geom_node_label(aes(col = splitvar),
                  line_list = list(aes(label = paste("Node", id)),
                                   aes(label = splitvar),
                                   aes(label = paste("p =", formatC(p.value, format = "e", digits = 2)))),
                  line_gpar = list(list(size = 12, col = "black", fontface = "bold"),
                                   list(size = 20),
                                   list(size = 12)),
                  ids = "inner") +
  geom_node_label(aes(label = paste0("Node ", id, ", N = ", nodesize)),
                  fontface = "bold",
                  ids = "terminal",
                  size = 5, 
                  nudge_y = 0.01) +
  theme(legend.position = "none")

```
  
## How to Use
**ggparty** tries to stick as closely as possible to **ggplot2**'s grammar of graphics. The basic building blocks to a **ggparty** plot are:  

* **ggparty()** replaces the usual `ggplot()`. Takes a object of class `'party'` instead of a `'data.frame'`.

* **geom_edge()** draws the edges between the nodes 

* **geom_edge_label()** labels the edges with the corresponding split breaks  

* **geom_node_label()** labels the nodes with the split variable, node info or anything
else. The shorthand versions of this geom **geom_node_splitvar()** and
**geom_node_info()** have the correct defaults to write the split variables in
the inner nodes resp. the info in the terminal nodes.  

* **geom_node_plot()** creates a custom ggplot at the location of the node  

Make sure to check out the vignettes for detailed information on the usage of **ggparty**.

[![Build Status](https://travis-ci.org/martin-borkovec/ggparty.svg?branch=master)](https://travis-ci.org/martin-borkovec/ggparty)


