---
title: "presentation"
author: "Martin Borkovec"
date: "2019 M03 27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, dpi=300)
```

ggparty aims to extend ggplot2 functionality to the partykit package. It provides the necessary tools
to create clearly structured and highly customizable visualizations for tree-objects of the class party.
  
  

```{r}
library(ggparty)
```

Loading the ggparty package will also load partykit and ggplot2 and thereby provide all necessary functions. Let's first recreate the example from the partykit vignette (link?) 
```{r}
data("WeatherPlay", package = "partykit")
sp_o <- partysplit(1L, index = 1:3)
sp_h <- partysplit(3L, breaks = 75)
sp_w <- partysplit(4L, index = 1:2)
pn <- partynode(1L, split = sp_o, kids = list(
  partynode(2L, split = sp_h, kids = list(
    partynode(3L, info = "yes"),
    partynode(4L, info = "no"))),
  partynode(5L, info = "yes"),
  partynode(6L, split = sp_w, kids = list(
    partynode(7L, info = "yes"),
    partynode(8L, info = "no")))))
py <- party(pn, WeatherPlay)
```
Every ggparty plot starts with a call to the eponymous ggparty function which requires an object of class party. In most cases we will probably want to draw at least edges, edge labels and node labels, so we will have to call the respective functions. The default mappings of geom_edge and and geom_edge_label ensure that lines between the related nodes are drawn and the corresponding splitbreaks are plotted at their centers.  
Since the text we want to print on the nodes differs depending on the kind of node, we will call geom_label twice. Once for the inner nodes, to plot the split variables and once for the terminal nodes to plot the info elements of the tree, which in this case contain the play decision.
```{r Weatherplay, fig.width = 7}
ggparty(py) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(aes(label = splitvar),
                  ids = "inner") +
  geom_node_label(aes(label = info),
                  ids = "terminal")
```

## ggparty()
As we have seen, ggparty takes a tree of class party and allows us to plot the contained information in the style of
the ggplot package. This is made possible by the internal function get_plot_data, to which said tree is passed and a dataframe containing the relevant information is returned.

This data.frame is the data argument of the call to ggplot with all the necessary settings and specifications. The information accessible this way includes the *id*, the *x* and *y* coordinates, the *parent* node's id, the number of *kids*, a list of the *breaks*, the *level* (root being 0), the *nodesize* and the parent's coordinates *x_parent* and *y_parent* of the nodes^[The words in italic represent the respective column names.]. The data contained in the nodes is stored as lists inside the dataframe and is comfortably accessible only through the geom_node_plot function which will be introduced below.  


```{r, results = "asis"}
pander::pandoc.table(ggparty:::get_plot_data(py)[,1:15])
```


Thanks to the ggplot structure we can now map different aspects of our plot to poperties of the nodes. Whether that's sensible in this case is a different question.

```{r, fig.width = 7}
ggparty(py) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(aes(label = splitvar,
                      col = factor(level),
                      size = nodesize),
                  ids = "inner") +
  geom_node_label(aes(label = info,
                      col = factor(level),
                      size = nodesize),
                  ids = "terminal")
```

We can create a horizontal tree simply by setting horizontal in ggparty to TRUE.
```{r, fig.width = 7}
ggparty(py, horizontal = TRUE) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(aes(label = splitvar),
                  ids = "inner") +
  geom_node_label(aes(label = info),
                  ids = "terminal")
```

Now let's take a look at a constparty object created from the same data.
```{r}
n1 <- partynode(id = 1L, split = sp_o, kids = lapply(2L:4L, partynode))
t2 <- party(n1,
            data = WeatherPlay,
            fitted = data.frame(
              "(fitted)" = fitted_node(n1, data = WeatherPlay),
              "(response)" = WeatherPlay$play,
              check.names = FALSE),
            terms = terms(play ~ ., data = WeatherPlay)
)
t2 <- as.constparty(t2)
```

```{r, fig.width = 7}
ggparty(t2,
        add_vars = list(play_prob = function(data, party) {
          mean(party$data$play == "yes")})) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(aes(label = splitvar),
                  ids = "inner") +
  geom_node_label(aes(label = paste0("Playprob.:\n",play_prob)),
                  ids = "terminal")
```

To visualize the distribution of the variable play we will use the geom_node_plot function. It allows us to show the *data* of each node in its separate plot. For this to work, we have to specify the argument gglist. Basically we have to provide a list of all the ggplot components we would add to a ggplot call on the data element of a node.  

```{r, fig.width = 2, fig.asp=0.8}
ggplot(t2[2]$data) +
  geom_bar(aes(x = "", fill = play),
           position = position_fill()) +
  xlab("play")
```

So if we were to use the above code to create the desired plot for one node, we can instead pass the two components added to the ggplot call as a list to gglist and geom_node_plot will create a version of it for every specified node (per default the terminal plots). Keep in mind, that since it's a list we need to use "," instead of "+" to combine the components. 
```{r, fig.asp=0.6, fig.width = 7}
ggparty(t2) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(aes(label = splitvar),
                  ids = "inner") +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = play),
                                       position = position_fill()),
                              xlab("play")))
```

Thanks to the flexibility of ggplot we are also very flexible in creating these node plots so that the barplot can be easily changed into a pie chart. The additional argument size can be set to "nodesize" which changes the size of the node plot relative to the number of observations in the respective node. 

```{r, fig.width = 7}
ggparty(t2) +
  geom_edge() +
  geom_edge_label() +
  geom_node_label(aes(label = splitvar),
                  ids = "inner") +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = play),
                                       position = position_fill()),
                              coord_polar("y"),
                              theme_void()),
                size = "nodesize")
```





## Layout

```{r}
## Boston housing data
data("BostonHousing", package = "mlbench")
BostonHousing <- transform(BostonHousing,
                           chas = factor(chas, levels = 0:1, labels = c("no", "yes")),
                           rad = factor(rad, ordered = TRUE))

## linear model tree
bh_tree <- lmtree(medv ~ log(lstat) + I(rm^2) | zn +
                    indus + chas + nox + age + dis + rad + tax + crim + b + ptratio,
                  data = BostonHousing, minsize = 40)

```
Let's take a look at ggparty's layout system with the help of this lmtree based on Boston housing data set from mlbench.

```{r, fig.width= 7, fig.asp=1}
bh_plot <- ggparty(bh_tree, terminal_space = 0.5) +
  geom_edge() +
  geom_edge_label() +
  geom_node_splitvar() +
  geom_node_plot(gglist = list(
    geom_point(aes(y = medv, x = `log(lstat)`, col = chas),
               alpha = 0.6)),
    height = 0.5) +
  geom_node_plot(gglist = list(
    geom_point(aes(y = medv, x = `I(rm^2)`, col = chas),
               alpha = 0.6)),
    height = 0.5,
    nudge_y = -0.25)

bh_plot
```
ggparty positions all the nodes within the unit square. For vertical trees the root is always at (0.5, 1), for horizontal ones it is at (0, 0.5). The argument terminal_size specifies how much room should be left for terminal plots. The default value depends on the depth of the supplied tree. The terminal nodes are placed at this value and in case labels are drawn, they are drawn there. In case plots are to be drawn their top borders are aligned to this value, i.e. the terminal plots are not centered. Therefore reducing the height of a terminal node shrinks it towards the top.  

So if we want to plot multiple plots per node we have to keep that in mind and can achieve this for example like this.  
The first geom_node_plot only takes the argument height = 0.5 which halves its size and effectively makes it
occupy only the upper half of the area it would normaly do. For the second geom_node_plot we also specify
the size to be 0.5 but additionaly we have to specify nudge_y. Since the terminal space is set to be 0.5,
we know that the first plot now spans from 0.5 to 0.25. So we want to move the line where to place the
second plot to 0.25, i.e. nudge it from 0.5 by -0.25.  
  
Changing the theme from the default theme_void to one were gridlines are drawn 
allows us to see the above desribed layout structure.


```{r, fig.width=7, fig.asp = 1}
bh_plot + theme_bw()
```

We can use this information to manually set the positions of nodes. To do this
we must pass a data.frame containing the columns id, x and y to the layout
argument of ggparty. 

```{r, fig.width= 7, fig.asp=1}
ggparty(bh_tree, terminal_space = 0.5,
        # x and y values need to be between 0 and 1
        layout = data.frame(id = c(1, 2),
                            x = c(0.7, 0.3),
                            y = c(1, 0.9))
        ) +
  geom_edge() +
  geom_edge_label() +
  geom_node_splitvar() +
  geom_node_plot(gglist = list(
    geom_point(aes(y = medv, x = `log(lstat)`, col = chas),
               alpha = 0.6)),
    height = 0.5) +
  geom_node_plot(gglist = list(
    geom_point(aes(y = medv, x = `I(rm^2)`, col = chas),
               alpha = 0.6)),
    height = 0.5,
    nudge_y = -0.25) + 
  theme_bw()

```


## Predictions

Next we are going to apply ggparty on a tree of class modelparty. As the name suggests
these objects are model-based trees, therefore we might want to be able to make some predictions.

First the tree is created using the partykit infrastructure.

```{r}
data("GBSG2", package = "TH.data")
GBSG2$time <- GBSG2$time/365

library("survival")
wbreg <- function(y, x, start = NULL, weights = NULL, offset = NULL, ...) {
  survreg(y ~ 0 + x, weights = weights, dist = "weibull", ...)
}


logLik.survreg <- function(object, ...)
  structure(object$loglik[2], df = sum(object$df), class = "logLik")

gbsg2_tree <- mob(Surv(time, cens) ~ horTh + pnodes | age + tsize +
                    tgrade + progrec + estrec + menostat, data = GBSG2,
                  fit = wbreg, control = mob_control(minsize = 80))
```


Once again we call the respective functions and generate the tree. Additionally 
we now set the predict_arg of geom_node_plot to a named list. This list contains a 
function specified by us to generate data for the newdata argument of a call to
predict and futher arguments passed to this call. This function has to be named
"newdata" and takes the *data* of each node consecutively as the only argument. The newdata is
then stored together with the *prediction* in *predict_data*. If we want to access
it in our gglist we have to specify the data argument to it. Since it's not a 
quoting function like aes we have to quote the corresponding component with 
expression otherwise it will be evaluated too soon and cause an error.  

So in this case we want to create a sequence over the range of the metric variable 
pnodes and combine it once with the first level of the binary variable horTh and 
once with the second and then generate predictions of the type "quantile" with p set
to 0.5. 

```{r, fig.asp = 0.8, fig.width=7}
# function to generate newdata for predictions
generate_newdata <- function(data){
  z <- data.frame(horTh = factor(rep(c("yes", "no"),
                                     each = length(data$pnodes))),
                  pnodes = rep(seq(from = min(data$pnodes),
                                   to = max(data$pnodes),
                                   length.out = length(data$pnodes)),
                               2)
  )
  z$x <- model.matrix(~ ., data = z)
  z}

ggparty(gbsg2_tree, terminal_space = 0.8, horizontal = TRUE) +
  geom_edge() +
  geom_node_label(aes(label = splitvar),
                  ids = "inner") +
  geom_edge_label() +
  geom_node_plot(
    gglist = list(geom_point(aes(y = `Surv(time, cens).time`,
                                 x = pnodes,
                                 col = horTh),
                             alpha = 0.6),
                  expression(
                    geom_line(data = predict_data,
                              aes(x = pnodes,
                                  y = prediction,
                                  col = horTh),
                              size = 1.2)
                  ),
                  theme_bw(),
                  ylab("Survival Time")
    ),
    predict_arg = list(newdata = generate_newdata,
                       type = "quantile",
                       p = 0.5)
  )

```
  
geom_node_label allows us to create mulitline labels and specifiy individual
graphical parameters for each line. To do this, we must not set label in aes,
but instead pass a list of aes to the argument line_list. The order of the list
is the same as the order in which the lines will be printend. Additionally we
have to pass a list to line_gpar. This list must be the same length as line_list
and contain separate named lists of graphical parameters. If we don't want to
change anything for a specific line, the respective list hast to be an empty
list. Mapping with the mapping argument of geom_node_label still works and
affects all lines and the border. The line specific graphical arguments

This may seem very convoluted, but keep in mind, that we only have to go
through this process if we want to addres the graphical parameters of specific
lines. If we want to plot multiple lines and are fine with uniform graphical
parameters, we can just set the label in the aes of the mapping.  


```{r}
data("TeachingRatings", package = "AER")
tr <- subset(TeachingRatings, credits == "more")

tr_tree <- lmtree(eval ~ beauty | minority + age + gender + division + native +
                    tenure, data = tr, weights = students, caseweights = FALSE)
```
```{r, fig.width= 7, fig.asp= 1}
ggparty(tr_tree,
        terminal_space = 0.5) +
  geom_edge(size = 1.5) +
  geom_edge_label(colour = "grey", size = 4) +
  geom_node_plot(gglist = list(geom_point(aes(x = fitted_values,
                                              y = residuals,
                                              col = tenure,
                                              shape = minority),
                                          alpha = 0.8),
                               geom_hline(yintercept = 0),
                               theme_bw(base_size = 10)),
                 scales = "free_x",
                 ids = "terminal",
                 shared_axis_labels = T
  ) +
  geom_node_label(aes(col = splitvar),
                  line_list = list(aes(label = paste("Node", id)),
                                   aes(label = splitvar),
                                   aes(label = paste("p =", formatC(p.value, format = "e", digits = 2)))),
                  line_gpar = list(list(size = 8, col = "black", fontface = "bold"),
                                   list(size = 12),
                                   list(size = 8)),
                  ids = "inner") +
  geom_node_label(aes(label = paste0("Node ", id, ", N = ", nodesize)),
                  fontface = "bold",
                  ids = "terminal",
                  size = 3,
                  nudge_y = 0.01) +
  theme(legend.position = "none")
```


If we want to extract additional elements of the tree we can do it by specifiying add_vars accordingly.
```{r, fig.asp = 1}
data("TeachingRatings", package = "AER")
tr <- subset(TeachingRatings, credits == "more")

tr_tree <- lmtree(eval ~ beauty | minority + age + gender + division + native +
                    tenure, data = tr, weights = students, caseweights = FALSE)

# generate colors
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
ggred <- gg_color_hue(2)[1]
ggteal <- gg_color_hue(2)[2]

# create dataframe with densities
df <- data.frame(x_dens = numeric(), y_dens = numeric(), id = numeric(), breaks = character())
for (ids in c(2, 5)) {
  x_dens <- density(tr_tree[ids]$data$age)$x
  y_dens <- density(tr_tree[ids]$data$age)$y
  breaks <- rep("left", length(x_dens))
  if (ids == 2) breaks[x_dens > 50] <- "right"
  if (ids == 5) breaks[x_dens > 40] <- "right"
  df <- rbind(df, data.frame(x_dens, y_dens, id = ids, breaks))
  }

# get the party started
ggparty(tr_tree, terminal_space = 0.4,
        add_vars = list(breaks = "$node$split$breaks"),
        layout = data.frame(id = c(1, 2, 5, 7),
                            x = c(0.35, 0.15, 0.7, 0.8),
                            y = c(0.95, 0.6, 0.8, 0.55))) +
  geom_edge(size = 1.2,
            col = c(NA, ggred, ggred, ggteal, ggteal, ggred, ggteal, ggred, ggteal),
            alpha = 1) +
  geom_node_plot(ids = c(2,5),
                 add_data = df,
                 gglist = list(
                   expression(geom_line(data = add_data,
                                        aes(x = x_dens,
                                            y = y_dens),
                                        show.legend = FALSE,
                                        alpha = 0.8)
                   ),
                   expression(geom_ribbon(data = add_data,
                                          aes(x = x_dens,
                                              ymin = 0,
                                              ymax = y_dens,
                                              fill = breaks),
                                          show.legend = FALSE,
                                          alpha = 0.8)),
                   xlab("age"),
                   theme_bw(),
                   theme(axis.title.y = element_blank())),
                 size = 1.5,
                 height = 0.5
  ) + 
  geom_node_plot(ids = 1,
                 gglist = list(expression(geom_bar(aes(x = gender, fill = gender),
                                                   show.legend = FALSE,
                                                   alpha = .8)),
                               theme_bw(),
                               theme(axis.title.y = element_blank())),
                 size = 1.5,
                 height = 0.5
  ) +
  geom_node_plot(ids = 7,
                 gglist = list(geom_bar(aes(x = division, fill = division),
                                        show.legend = FALSE,
                                        alpha = .8),
                               theme_bw(),
                               theme(axis.title.y = element_blank())),
                 size = 1.5,
                 height = 0.5
  ) +
  geom_node_plot(gglist = list(geom_point(aes(x = beauty,
                                              y = eval,
                                              col = tenure,
                                              shape = minority),
                                          alpha = 0.8),
                               expression(
                                 geom_line(data = predict_data,
                                           aes(x = beauty,
                                               y = prediction),
                                           col = "blue")),
                               theme_bw(base_size = 10),
                               scale_color_manual(values = c(gg_color_hue(4)[2],
                                                             gg_color_hue(4)[4]))),
                 scales = "fixed",
                 ids = "terminal",
                 shared_axis_labels = T,
                 predict_arg = list(newdata = function(x) {
                   data.frame(beauty = seq(min(x$beauty),
                                           max(x$beauty),
                                           length.out = 100))
                 })) +
  theme(legend.position = "none")
```

```{r}
autoplot(py)
```

```{r}
autoplot(t2)
```

```{r, fig.asp = 1}
autoplot(bh_tree, plot_var = "log(lstat)", show_fit = FALSE)
autoplot(bh_tree, plot_var = "I(rm^2)", show_fit = TRUE)
```

```{r}
autoplot(gbsg2_tree, plot_var = "pnodes")
```

```{r, fig.asp = 1}
autoplot(tr_tree)
```




